---
title: "Live Coding - Data Science Lifecycle (Solutions)"
format: html
editor: visual
execute:
  echo: true
  warning: false
editor_options:
  chunk_output_type: console
---

```{r}
library(gapminder) # R package that contains the data
library(dplyr)
library(ggplot2)
library(knitr)
```

# Data import

Data can be imported from many different sources. In this exercise, we import data from an R package loaded via `library()`.

## Gapminder data

The data is an [excerpt of the Gapminder data](https://www.gapminder.org/data/documentation/) on life expectancy, GDP per capita, and population by country, compiled for teaching purposes (up to 2007).

```{r}
gapminder
```

# Data tidying

The gapminder data is already tidy: each row is one country-year observation and each column is a single variable, so no tidying step is needed before transformation.

## Gapminder data

```{r}
glimpse(gapminder)
```

# Data transformation

## Gapminder data

**Goal:** Calculate the median life expectancy at birth by continent for 2007.

Step 1: Filter the data to only include the year 2007.

```{r}
gapminder_yr_2007 <- gapminder |>
  filter(year == 2007)
```

Step 2: Group the data by continent and calculate the median life expectancy.

```{r}
gapminder_summary_continent <- gapminder_yr_2007 |>
  group_by(continent) |>
  summarise(lifeExp = median(lifeExp))
```

# Data visualisation (and tables)

## Gapminder data

Africa has the lowest life expectancy and the greatest spread in life expectancy globally.

Boxplot with jittered points showing the distribution of life expectancy by continent in 2007:

```{r}
ggplot(data = gapminder_yr_2007,
       mapping = aes(x = continent,
                     y = lifeExp)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 1/4, size = 3) +
  labs(x = NULL,
       y = "life expectancy") +
  theme_minimal(base_size = 14)
```

Summary table of median life expectancy by continent in 2007:

```{r}
kable(gapminder_summary_continent,
      col.names = c("Continent", "Median life expectancy (2007)"),
      digits = 1)
```

# Data communication

Your rendered html file is a data communication product. You can share it directly, host it as a website, or render to pdf/docx.

# Stage, Commit & Push to GitHub

Workflow only. Nothing to code here.
